---
title: "Diversity - EF bootstrapped"
author: "Fabian Roger"
date: "28 Nov 2015"
output:
  html_document:
    fig_caption: yes
    toc: yes
  pdf_document:
    latex_engine: lualatex
    toc: yes
---

In this document we will correlate the three metrics of diveristy with the three ecosystem function we measured:

Diveristy metrics:

+ Species richness (effective number of species - Hill1)
+ Phylogenetic diveristy (Effective number of taxonomicaly completely distinct species)
+ Functional diveristy (effectiv numer of carbon sources that a community can consume)

Ecosystem functions:

+ Maximum Cell yield (averaged over the five highest cell yield sobserved for each community)
+ Stability (inverse of the coefficent of variation of the detrended cell yield over the course of the experiment)
+ remaining dissolev organic nitrogen at the end of the experiment. 

for that we first need to import the according data

**Ecosystem Functions**
```{r}
maxBM <- read.table("response_var/maxBM.txt", header = TRUE, stringsAsFactors = FALSE)
Stability <- read.table("response_var/Stability.txt", header = TRUE, stringsAsFactors = FALSE)
NUT <- read.table("response_var/NUT.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)
```

**Diversity metrics**
```{r}
DIV <- read.table("OTU_docs/DIV.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE)
FunDiv <- read.table("EcoLogDocs/FunDiv.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE)
```

we can join the functional diveristy data to the rest and plot them, both against dilution and against each other

```{r, echo=FALSE, warning=FALSE}
DIV <- na.omit(DIV) %>% 
  left_join( FunDiv) %>% 
  transform(Sampling = factor(Sampling, levels = c("14.06", "28.06", "12.07")))

ggpairs(DIV,
        columns = 5:7, 
        colour = 'Sampling', 
        columnLabels = c("effective number\nof species", "phylogenetic\ndiversity", "functional\ndiveristy"),
        lower = list(continuous = "smooth"), 
        title = "correlation between the diveristy metrics by sampling date")+
  theme_bw(base_size = 10) 
```

```{r, echo=FALSE, warning=FALSE}

ggplot(DIV, aes(x = DIL, y = effN, colour = Lake, shape = Lake, group = Lake))+
  geom_point()+
  facet_wrap(Sampling~Lake)+
  stat_smooth(method = "lm", se = F, linetype = "dashed")+
  scale_y_log10(breaks = c(2, 4, 8, 16, 32))+
  theme_bw()+
  scale_colour_manual(values = c("orange", "darkred", "darkgreen", "darkblue"  ))+
  labs(title = "species richness",
       y = "effective number of species (q = 1)",
       x = "dillution factor")+
  theme(legend.position = "none",
        axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5))
  

ggplot(DIV, aes(x = DIL, y = PD, colour = Lake, shape = Lake, group = Lake))+
  geom_point()+
  facet_wrap(Sampling~Lake)+
  stat_smooth(method = "lm", se = F, linetype = "dashed")+
  scale_y_log10(breaks = c(2, 4, 8, 16))+
  theme_bw()+
  scale_colour_manual(values = c("orange", "darkred", "darkgreen", "darkblue"  ))+
  labs(title = "phylogenetic diversity (q = 1)",
       y = "effective number of phylogenetic disctinct species",
       x = "dillution factor")+
  theme(legend.position = "none",
        axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5))

ggplot(DIV, aes(x = DIL, y = FuncDiv , colour = Lake, shape = Lake, group = Lake))+
  geom_point()+
  facet_wrap(Sampling~Lake)+
  stat_smooth(method = "lm", se = F, linetype = "dashed")+
  theme_bw()+
  scale_colour_manual(values = c("orange", "darkred", "darkgreen", "darkblue"  ))+
  labs(title = "functional diveristy",
       y = "effective number of utilized carbon sources",
       x = "dillution factor")+
  theme(legend.position = "none",
        axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5))

```


For each diveristy metric, we have three measurements, taken at three timepoints. Our measure of ecosystem funcitoning however result in a single value per dilution level and Lake, as they integrate of the whole course of the experiment. 

Therefore we have to take the average diveristy as predictor variable. In the graph below we show the value of each diverist metric plot together for the three sampling dates (coloured dots) and the average diversity (black dots)

```{r, echo = FALSE, warning = FALSE}
G1 <- DIV %>% 
  melt(id.var = c("BOT", "Sampling", "Lake", "DIL")) %>%
  group_by(BOT, Lake, DIL, variable) %>% 
  dplyr::mutate(y = mean(value)) %>% 
  ggplot(aes(x = DIL, y = y, colour = Lake, shape = Lake, group = 1))+
  geom_point(aes(x = DIL, y = value), alpha = 0.4, size = 1.5)+
  geom_point(size = 2)+
  facet_grid(variable ~ Lake, scales = "free_y")+
  stat_smooth(method = "lm", se = F, linetype = "dashed")+
  scale_colour_manual(values = c("orange", "darkred", "darkgreen", "darkblue"  ))+
  scale_y_log10(breaks = c(2,4,8,16,32))+
  theme_bw()+
  labs(title = "average diversity over the three smapling dates",
       y = "value of diversiyt metric",
       x = "dillution factor")+
  theme(legend.position = "none",
        axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5))


```


```{r}
ggsave(plot = G1, "figures/Figure_1.pdf", height = 8, width = 8)
```


```{r}
avDIV <- dplyr::select(DIV, -Sampling) %>% 
  group_by(BOT, Lake, DIL) %>% 
  summarise_each(funs(mean(., na.rm = T)))
```

now we can correlate each diveristy metric to each ecosystem function for each Lake. In order to be able to comapre the absolut value of the slopes, we scale all varibales to mean = 0 and sd = 1 prior to corrleation. Note that the nutrient data is already scaled so we join it after having scaled the other variables. 

Also, before we proceed, we exclude dilution level 0 form the data. See the manuscript for a discussion why

 
```{r}

BEF <- filter(avDIV, DIL != 0) %>% 
  left_join(maxBM) %>% 
  left_join(Stability) %>% 
  group_by(Lake) %>% 
  left_join(NUT)

head(BEF,2) 
```

as a first attmept we will produce bivariate scatterplots 

```{r}

BEF_long <- BEF %>% 
  ungroup %>% 
  gather(DIVmetric, DIVval, effN, PD, FuncDiv)


BEF_long2 <- BEF %>% 
  ungroup %>% 
  gather(DIVmetric, DIVval, effN, PD, FuncDiv) %>% 
  gather(EFmetric, EFval, maxCells, Stability, DIN) %>% 
  dplyr::select(-BOT, -DIL)

LM <- BEF_long2 %>%
  group_by(Lake, DIVmetric, EFmetric) %>% 
  do( glance( lm( scale(EFval) ~ scale(DIVval), .))) %>% 
  dplyr::select(Lake, DIVmetric, EFmetric, r.squared, p.value)

Label_x <- BEF_long2 %>%
  group_by(DIVmetric, EFmetric) %>% 
  dplyr::summarize(x = (max(DIVval) - (max( DIVval) - min(DIVval))* 0.2 ))

Label_y <- BEF_long2 %>%
  group_by(Lake, EFmetric) %>% 
  dplyr::summarize(y1 = (max(EFval) - (max( EFval) - min(EFval))* 0.2 ),
                   y2 = (max(EFval) - (max( EFval) - min(EFval))* 0.35 ))


LM <- left_join(LM, Label_x) %>%  left_join(Label_y)
  



scientific_10 <- function(x) {
  parse(text = gsub("e", " %*% 10^", scientific_format()(x)))
}

LMmC <- filter(LM, EFmetric == "maxCells")
  
G2 <-  ggplot(BEF_long, aes(x = DIVval, y = maxCells, colour = Lake, shape = Lake))+
  geom_point()+
  geom_text(data = LMmC, aes(x = x, y = y1, label = paste("r2.", signif(r.squared, 1), sep = " : " )), colour = "black", size = 2)+
  geom_text(data = LMmC, aes(x = x, y = y2, label = paste("p", signif(p.value, 1), sep = " : " )), colour = "black", size = 2)+
  facet_grid(Lake ~ DIVmetric, scales = "free")+
  theme_bw()+
  stat_smooth(method = "lm", se = FALSE, linetype = "dashed")+
  labs(title = "maximum cell yield",
       y = "maximum Cell yield",
       x = "value of diveristy metric")+
  theme(legend.position = "none",
        axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5))+
  scale_y_continuous(label=scientific_10)+
  scale_colour_manual(values = c("orange", "darkred", "darkgreen", "darkblue"  ))


LM_S <- filter(LM, EFmetric == "Stability")

G3 <- ggplot(BEF_long, aes(x = DIVval, y = Stability, colour = Lake))+
    geom_point()+
  geom_text(data = LM_S, aes(x = x, y = y1, label = paste("r2.", signif(r.squared, 1), sep = " : " )), colour = "black", size = 2)+
  geom_text(data = LM_S, aes(x = x, y = y2, label = paste("p", signif(p.value, 1), sep = " : " )), colour = "black", size = 2)+
    facet_grid(Lake ~ DIVmetric, scales = "free")+
    theme_bw()+
    stat_smooth(method = "lm", se = FALSE, linetype = "dashed")+
    labs(title = "temporal stability of cell yield",
         y = "stability of cell yield",
         x = "value of diveristy metric")+
    theme(legend.position = "none",
          axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5))+
    scale_colour_manual(values = c("orange", "darkred", "darkgreen", "darkblue"  ))

LM_DIN <- filter(LM, EFmetric == "DIN")
  
G4 <- ggplot(BEF_long, aes(x = DIVval, y = DIN, colour = Lake))+
    geom_point()+
  geom_text(data = LM_DIN, aes(x = x, y = y1, label = paste("r2.", signif(r.squared, 1), sep = " : " )), colour = "black", size = 2)+
  geom_text(data = LM_DIN, aes(x = x, y = y2, label = paste("p", signif(p.value, 1), sep = " : " )), colour = "black", size = 2)+
    facet_grid(Lake ~ DIVmetric, scales = "free")+
    theme_bw()+
    stat_smooth(method = "lm", se = FALSE, linetype = "dashed")+
    labs(title = "depletion of dissolved organic nitrogene",
         y = "-1 * standardized DIN concentration",
         x = "value of diveristy metric")+
    theme(legend.position = "none",
          axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5))+
    scale_colour_manual(values = c("orange", "darkred", "darkgreen", "darkblue"  ))

  
MF_DF <- BEF_long %>% 
  group_by(DIVmetric, Lake) %>% 
  do( getFuncMaxed(., vars = c("DIN", "Stability", "maxCells"), thresh = 0.75, prepend = "DIVval", maxN = 2))

LM_MF <- MF_DF %>%
  group_by(Lake, DIVmetric) %>% 
  do( glance( lm(scale(funcMaxed) ~ scale(DIVval), .))) %>% 
  dplyr::select(Lake, DIVmetric, r.squared, p.value)

LM_MF <- Label_x %>% dplyr::select(-EFmetric) %>% distinct %>% left_join(LM_MF ,.)



G5 <- ggplot(MF_DF, aes(x = DIVval, y = funcMaxed, colour = Lake))+
    geom_point()+
  geom_text(data = LM_MF, aes(x = x, y = 2.5, label = paste("r2.", signif(r.squared, 1), sep = " : " )), colour = "black", size = 2)+
  geom_text(data = LM_MF, aes(x = x, y = 2, label = paste("p", signif(p.value, 1), sep = " : " )), colour = "black", size = 2)+
    facet_grid(Lake ~ DIVmetric, scales="free_x")+
    theme_bw()+
    stat_smooth(method = "lm", se = FALSE, linetype = "dashed")+
    labs(title = "multifunctionality at 75 % threshold",
         y = "number of function > 75% of maximum",
         x = "value of diveristy metric")+
    theme(legend.position = "none",
          axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5))+
    scale_colour_manual(values = c("orange", "darkred", "darkgreen", "darkblue"  ))+
  theme(text=element_text(family="times"))
  
  
Div_G <- arrangeGrob(G2, G3, G4, G5, ncol = 2)
Div_G
ggsave(plot = Div_G, "figures/Figure_3.pdf", height = 10, width = 10)
  
  

```

It is visible from the scatterplots that there doesn't seem to be a strong correlation between either of the diveristy metrics and either of the metrics of ecosystem functioning. 

However, it is also visible that the data are not ideally distributed or at least not in all cases, wherfore we will use bootstrap as robust test of a positive relationship. To bring the slopes on the same scale, we will scale all variables by their mean and standard deviation. `DIN` is allready scaled wherefore it is not scaled again. 

```{r}

scale_numeric <- function(x) {(x- mean(x)) / sd(x)} 

BEF_scaled <- BEF %>% 
  group_by(Lake ) %>% 
  mutate_each(funs(scale_numeric), -BOT, -DIL, -DIN)

BEF_scaled_long <- BEF_scaled %>% 
  ungroup %>% 
  gather(DIVmetric, DIVval, effN, PD, FuncDiv) %>% 
  gather(EFmetric, EFval, maxCells, Stability, DIN) %>% 
  dplyr::select(-BOT, -DIL) 

BEF_list <- split(BEF_scaled_long, list(BEF_scaled_long$Lake,
                            BEF_scaled_long$DIVmetric,
                            BEF_scaled_long$EFmetric))



Boot_LM <- function(x,N) {x %>% 
    bootstrap(N) %>%
    do( tidy( lm( EFval ~ DIVval, .)))}

N <- 100 #number of bootstrapes

Slopes <- mclapply(BEF_list, Boot_LM, N=N)# bootstrap each Lake * Div_metric * EF_metric combination
Slopes_Df <- do.call("rbind", Slopes) # rbin all dataframes in list

# add colum with names of dataframe (Lake.Div_metric.EF_metric), N replicates, 2 rows per replicate (intercept and slope)
Slopes_Df$var <- rep(names(Slopes), each = N*2) 

Slopes_Df <- Slopes_Df %>% 
  separate(var, c("Lake", "DIVmetric", "EFmetric"), sep = "\\.") %>% 
  filter(term == "DIVval")

write.table(Slopes_Df, "boot_Slopes_10k.txt", sep = "\t")

prct <- function(x) (length(x[x > 0]) / length(x)) * 100

Slopes_prct <- Slopes_Df %>% 
  group_by(Lake, DIVmetric, EFmetric) %>% 
  summarise(prct_pos = round( prct(estimate), 1)) %>% 
  mutate(prct_neg = round((100 - prct_pos), 1))
  

G7 <- Slopes_Df %>% 
  filter(EFmetric == "maxCells") %>% 
  ggplot(aes(x = estimate, fill = Lake))+
  geom_density(binwidth = 0.01)+
  #geom_text(data = Slopes_prct[Slopes_prct$EFmetric == "maxCells",],
   #         aes(x = 1.2, y = 500, label = paste(prct_pos, "%", sep = " "), size = 7))+
  #geom_text(data = Slopes_prct[Slopes_prct$EFmetric == "maxCells",],
  #          aes(x = - 1.2, y = 500, label = paste(prct_neg, "%", sep = " "), size = 7))+
  facet_grid(DIVmetric~Lake)+
  geom_vline(xintercept = 0, linetype = "dashed")+
  theme_bw()+
  scale_fill_manual(values = c("orange", "darkred", "darkgreen", "darkblue"  ))+
 # scale_y_continuous(limits = c(0,600))+
  scale_x_continuous(limits = c(-2,2))+
  labs(title = "maximum cell yield", x = "")+
  theme(legend.position = "none")


G8 <- Slopes_Df %>% 
  filter(EFmetric == "Stability") %>% 
  ggplot(aes(x = estimate, fill = Lake))+
  geom_density(binwidth = 0.01)+
  #geom_text(data = Slopes_prct[Slopes_prct$EFmetric == "Stability",],
   #         aes(x = 1.2, y = 500, label = paste(prct_pos, "%", sep = " "), size = 7))+
  #geom_text(data = Slopes_prct[Slopes_prct$EFmetric == "Stability",],
   #         aes(x = - 1.2, y = 500, label = paste(prct_neg, "%", sep = " "), size = 7))+
  facet_grid(DIVmetric~Lake)+
  geom_vline(xintercept = 0, linetype = "dashed")+
  theme_bw()+
  scale_fill_manual(values = c("orange", "darkred", "darkgreen", "darkblue"  ))+
  #scale_y_continuous(limits = c(0,600))+
  scale_x_continuous(limits = c(-2,2))+
  labs(title = "stability ", x = "", y = "")+
  theme(legend.position = "none")

G9 <- Slopes_Df %>% 
  filter(EFmetric == "DIN") %>% 
  ggplot(aes(x = estimate, fill = Lake))+
  geom_density( binwidth = 0.01)+
  #geom_text(data = Slopes_prct[Slopes_prct$EFmetric == "DIN",],
   #         aes(x = 1.2, y = 500, label = paste(prct_pos, "%", sep = " "), size = 7))+
  #geom_text(data = Slopes_prct[Slopes_prct$EFmetric == "DIN",],
   #         aes(x = - 1.2, y = 500, label = paste(prct_neg, "%", sep = " "), size = 7))+
  geom_vline(xintercept = 0, linetype = "dashed")+
  facet_grid(DIVmetric~Lake)+
  theme_bw()+
  scale_fill_manual(values = c("orange", "darkred", "darkgreen", "darkblue"  ))+
 # scale_y_continuous(limits = c(0,600))+
  scale_x_continuous(limits = c(-2,2))+
  labs(title = "dissolved inorganic nitrogen ", x = "bootstraped slope estimate")+
  theme(legend.position = "none")
  


```


```{r}
# multifunctionality

BEF_scaled_MF <- BEF %>% 
  ungroup %>% 
  gather(DIVmetric, DIVval, effN, PD, FuncDiv) %>% 
  dplyr::select(-BOT, -DIL) 

BEF_list_MF <- split(BEF_scaled_MF, list(BEF_scaled_MF$Lake,
                            BEF_scaled_MF$DIVmetric))

BEF_list_MF$Surtesjön.effN

Boot_MF <- function(x, N, thresh, maxN) {
  x %>% 
  bootstrap(N) %>%
    do( tidy( lm( funcMaxed ~ DIVval, 
                  data = getFuncMaxed(.,
                                      vars = c("DIN", "Stability", "maxCells"),
                                      thresh = thresh, 
                                      prepend = "DIVval", 
                                      maxN = maxN))))
      }

N = 10000

MF_Slopes <- mclapply(BEF_list_MF, Boot_MF, N = N, thresh = 0.75, maxN = 2)

MF_Slopes_Df <- do.call("rbind", MF_Slopes) # rbin all dataframes in list

# add colum with names of dataframe (Lake.Div_metric.EF_metric), N replicates, 2 rows per replicate (intercept and slope)
MF_Slopes_Df$var <- rep(names(MF_Slopes), each = N*2) 

MF_Slopes_Df <- MF_Slopes_Df %>% 
  separate(var, c("Lake", "DIVmetric"), sep = "\\.") %>% 
  filter(term == "DIVval")

write.table(MF_Slopes_Df, "boot_MF_Slopes_10k.txt", sep = "\t")

MF_Slopes_prct <- MF_Slopes_Df %>% 
  group_by(Lake, DIVmetric) %>% 
  dplyr::summarise(prct_pos = round( prct(estimate), 1)) %>% 
  mutate(prct_neg = round((100 - prct_pos), 1))


G10 <- ggplot(MF_Slopes_Df, aes(x = estimate, fill = Lake))+
  geom_histogram( binwidth = 0.005)+
  geom_text(data = MF_Slopes_prct,
            aes(x = 1.2, y = 500, label = paste(prct_pos, "%", sep = " "), size = 7))+
  geom_text(data = MF_Slopes_prct,
            aes(x = - 1.2, y = 500, label = paste(prct_neg, "%", sep = " "), size = 7))+
  geom_vline(xintercept = 0, linetype = "dashed")+
  facet_grid(DIVmetric~Lake)+
  theme_bw()+
  scale_fill_manual(values = c("orange", "darkred", "darkgreen", "darkblue"  ))+
  #scale_y_continuous(limits = c(0,600))+
  scale_x_continuous(limits = c(-2,2))+
  labs(title = "multifunctionality at 75% of maximum ", x = "bootstraped slope estimate", y="")+
  theme(legend.position = "none")

Slope_Graph <- arrangeGrob(G7, G8, G9, G10, ncol = 2)
Slope_Graph
ggsave(plot = Slope_Graph, "figures/Figure_3_dens.pdf", height = 12, width = 15)
  

```



```{r}

library(dplyr)
library(tidyr)

DF <- data.frame(A=rnorm(100, 100, 10))
DF$B <- DF$A + rnorm(100, 0, 1)
DF$C <- DF$A + rnorm(100, 0, 5)

G1 <- ggplot(DF, aes(x = A, y = B))+
  geom_point()

G2 <- ggplot(DF, aes(x = A, y = C))+
  geom_point()


S1 <- DF %>% 
  bootstrap(1000) %>%
    do( tidy( lm( A ~ B, data = .))) %>% 
  filter(term == "B")

quantile(S1$estimate, c(0.025, 0.5, 0.975))
confint(lm(A~B, data = DF), "C", level = 0.95)

G3 <- ggplot(S1, aes(x = estimate))+
  geom_density(binwidth = 0.001)+
  scale_x_continuous(limits = c(0,1.5))

S2 <- DF %>% 
  bootstrap(1000) %>%
    do( tidy( lm( A ~ C, data = .))) %>% 
  filter(term == "C") 

quantile(S2$estimate, c(0.025, 0.5, 0.975))
confint(lm(A~C, data = DF), "C", level = 0.95)

G5 <- ggplot( S2, aes(x = estimate))+
  geom_density(binwidth = 0.001)+
  scale_x_continuous(limits = c(-1.5,1.5))

grid.arrange(G1,G2,G3,G4)



```


