---
title: "Diversity - EF bootstrapped"
author: "Fabian Roger"
date: "28 Nov 2015"
output:
  html_document:
    fig_caption: yes
    toc: yes
  pdf_document:
    latex_engine: lualatex
    toc: yes
---

__this script imports:__ 

+ response_var/maxBM.txt        *maximum cell counts*                 
+ response_var/Stability.txt    *temporal stability of the cell counts*
+ response_var/NUT.txt          *rescaled dissolved inorganic nitrogen data, high values = high depletion*
+ OTU_docs/DIV.txt              *effective number of species and phylogenetic diversity based on hill numbers*
+ EcoLogDocs/FunDiv.txt         *functional diversity based on Biolog EcoPlates*
   
__this script does:__

+ joins the diveristy metrics and measured ecosystem functions
+ calculates the average diveristy 
+ calcuates multifunctionality
+ correlates the diveristy metrics with the ecosystem funcitons
+ coputes r squares and p-values for each correlation
  
__this script exports:__
  
figures:
  
+ figures/Figure_3.pdf      *biplot of diveristy *
  

```{r, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message=FALSE)
```
```{r}
source("utility_functions.R")
```
```{r, echo=FALSE}
require(dplyr)
require(GGally)
require(ggplot2)
require(tidyr)
require(broom)
require(scales)
require(gridExtra)
require(grid)
```

###The Metrics

####Diveristy metrics:

+ Species richness (effective number of species - Hill1)
+ Phylogenetic diveristy (Effective number of taxonomicaly completely distinct species)
+ Functional diveristy (effectiv numer of carbon sources that a community can consume)

####Ecosystem functions:

+ Maximum Cell yield (averaged over the five highest cell yield sobserved for each community)
+ Stability (inverse of the coefficent of variation of the detrended cell yield over the course of the experiment)
+ remaining dissolev organic nitrogen at the end of the experiment. 

for that we first need to import the according data

**Ecosystem Functions**
```{r}
maxBM <- read.table("response_var/maxBM.txt", header = TRUE, stringsAsFactors = FALSE)
Stability <- read.table("response_var/Stability.txt", header = TRUE, stringsAsFactors = FALSE)
NUT <- read.table("response_var/NUT.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)
```

**Diversity metrics**
```{r}
DIV <- read.table("OTU_docs/DIV.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE)
FunDiv <- read.table("EcoLogDocs/FunDiv.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE)
```

we can join the functional diveristy data to the rest and plot them, both against dilution and against each other

```{r, echo=FALSE}
DIV <- na.omit(DIV) %>% 
  left_join( FunDiv) %>% 
  transform(Sampling = factor(Sampling, levels = c("14.06", "28.06", "12.07")))

divPairs2 <- ggpairs(DIV,
        columns = 5:7, 
        colour = 'Sampling', 
        columnLabels = c("effective number\nof species", "phylogenetic\ndiversity", "functional\ndiveristy"),
        lower = list(continuous = "smooth"), 
        title = "correlation between the diveristy metrics by sampling date")+
  theme_bw(base_size = 10) 

divPairs2

pdf("figures/Fig_maybe_3.pdf")
print(divPairs2)
dev.off()


```

```{r, echo=FALSE}

ggplot(DIV, aes(x = DIL, y = effN, colour = Lake, shape = Lake, group = Lake))+
  geom_point()+
  facet_wrap(Sampling~Lake)+
  stat_smooth(method = "lm", se = F, linetype = "dashed")+
  scale_y_log10(breaks = c(2, 4, 8, 16, 32))+
  theme_bw()+
  scale_colour_manual(values = c("orange", "darkred", "darkgreen", "darkblue"  ))+
  labs(title = "species richness",
       y = "effective number of species (q = 1)",
       x = "dillution factor")+
  theme(legend.position = "none",
        axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5))
  

ggplot(DIV, aes(x = DIL, y = PD, colour = Lake, shape = Lake, group = Lake))+
  geom_point()+
  facet_wrap(Sampling~Lake)+
  stat_smooth(method = "lm", se = F, linetype = "dashed")+
  scale_y_log10(breaks = c(2, 4, 8, 16))+
  theme_bw()+
  scale_colour_manual(values = c("orange", "darkred", "darkgreen", "darkblue"  ))+
  labs(title = "phylogenetic diversity (q = 1)",
       y = "effective number of phylogenetic disctinct species",
       x = "dillution factor")+
  theme(legend.position = "none",
        axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5))

ggplot(DIV, aes(x = DIL, y = FuncDiv , colour = Lake, shape = Lake, group = Lake))+
  geom_point()+
  facet_wrap(Sampling~Lake)+
  stat_smooth(method = "lm", se = F, linetype = "dashed")+
  theme_bw()+
  scale_colour_manual(values = c("orange", "darkred", "darkgreen", "darkblue"  ))+
  labs(title = "functional diveristy",
       y = "effective number of utilized carbon sources",
       x = "dillution factor")+
  theme(legend.position = "none",
        axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5))

```


For each diveristy metric, we have three measurements, taken at three timepoints. Our measure of ecosystem funcitoning however result in a single value per dilution level and Lake, as they integrate the whole course of the experiment. 

Therefore we have to take the average diveristy as predictor variable. In the graph below we show the value of each diverist metric plot together for the three sampling dates (coloured dots) and the average diversity (black dots)

```{r, echo = FALSE}
G1 <- DIV %>% 
  gather(DIVmetric, DIVvalue, effN, PD, FuncDiv) %>%
  group_by(BOT, Lake, DIL, DIVmetric) %>% 
  mutate(y = mean(DIVvalue)) %>% 
  ggplot(aes(x = DIL, y = y, shape = Lake, group = 1))+
  geom_point(aes(x = DIL, y = DIVvalue), alpha = 0.3, size = 1.5)+
  geom_point(size = 2)+
  facet_grid(DIVmetric ~ Lake, scales = "free_y")+
  stat_smooth(method = "lm", se = F, linetype = "solid", colour = "black", size = 0.5)+
  scale_y_log10(breaks = c(2,4,8,16,32))+
  theme_bw()+
  labs(title = "average diversity over the three smapling dates",
       y = "value of diversity metric",
       x = "dillution factor")+
  theme(legend.position = "none",
        axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5))

G1
```


```{r}
ggsave(plot = G1, "figures/Figure_1.pdf", height = 8, width = 8)
```


```{r}
avDIV <- select(DIV, -Sampling) %>% 
  group_by(BOT, Lake, DIL) %>% 
  summarise_each(funs(mean(., na.rm = T)))
```

now we can correlate each diveristy metric to each ecosystem function for each Lake. In order to be able to comapre the absolut value of the slopes, we scale all varibales to mean = 0 and sd = 1 prior to corrleation. Note that the nutrient data is already scaled so we join it after having scaled the other variables. 

Also, before we proceed, we exclude dilution level 0 form the data. See the manuscript for a discussion why

 
```{r}

BEF <- filter(avDIV, DIL != 0) %>% 
  left_join(maxBM) %>% 
  left_join(Stability) %>% 
  group_by(Lake) %>% 
  left_join(NUT)

head(BEF,2) 
```

To see the correlations between the three metrics of EF and the three diverist metrics, we correlate them 

```{r, echo = FALSE}

BEF_long <- BEF %>%  # gather diveristy metrics in one colum, keep EF separate
  ungroup %>% 
  gather(DIVmetric, DIVval, effN, PD, FuncDiv)


BEF_long2 <- BEF %>% #gather both diveristy columns and EF in a single column respectively
  ungroup %>% 
  gather(DIVmetric, DIVval, effN, PD, FuncDiv) %>% 
  gather(EFmetric, EFval, maxCells, Stability, DIN) %>% 
  select(-BOT, -DIL)

LM <- BEF_long2 %>% #regress each div metric against each EF, by Lake. extract p-value and r-square
  group_by(Lake, DIVmetric, EFmetric) %>% 
  do( glance( lm( scale(EFval) ~ scale(DIVval), .))) %>% 
  select(Lake, DIVmetric, EFmetric, r.squared, p.value)

Label_x <- BEF_long2 %>% #calculate x value for where to add the linear regression statistic in respective panels
  group_by(DIVmetric, EFmetric) %>% 
  summarize(x = (max(DIVval) - (max( DIVval) - min(DIVval))* 0.2 ))

Label_y <- BEF_long2 %>% #calculate y values to print the add regression statistics in respective panels
  group_by(Lake, EFmetric) %>% 
  summarize(y1 = (max(EFval) - (max( EFval) - min(EFval))* 0.2 ),
                   y2 = (max(EFval) - (max( EFval) - min(EFval))* 0.35 ))

Label_ymin_ymax <- BEF_long2 %>% #calculate ymin value for where to beginn the white transparent rectangle behind teh text
  group_by(Lake, EFmetric) %>% 
  summarize(ymax = (max(EFval) - (max( EFval) - min(EFval))* 0.15 ),
                   ymin = (max(EFval) - (max( EFval) - min(EFval))* 0.4 ))


LM <- left_join(LM, Label_x) %>%  left_join(Label_y) %>% left_join(Label_ymin_ymax)

LM$xmin <- LM$x - 0.2 * LM$x
LM$xmax <- LM$x + 0.2 * LM$x


  
LMmC <- filter(LM, EFmetric == "maxCells") # subset for and plot EF: maxCells
  
G2 <-  ggplot(BEF_long, aes(x = DIVval, y = maxCells, shape = Lake))+
  geom_point()+
  stat_smooth(method = "lm", se = FALSE, linetype = "solid", size = 0.5, colour = "black")+
  geom_rect( data = LMmC, aes(x=NULL,y=NULL, xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
             fill = "white", alpha = 0.8) +
  geom_text(data = LMmC, aes(x = x, y = y1, label = paste("r2.", signif(r.squared, 1), sep = " : " )), colour = "black", size = 2)+
  geom_text(data = LMmC, aes(x = x, y = y2, label = paste("p", signif(p.value, 1), sep = " : " )), colour = "black", size = 2)+
  facet_grid(Lake ~ DIVmetric, scales = "free")+
  theme_bw()+
  labs(title = "maximum cell yield",
       y = "maximum Cell yield",
       x = "")+
  theme(legend.position = "none",
       axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5),
       plot.margin = unit(c(1,0.5,0,1), "cm"))+
  scale_y_continuous(label=scientific_10)# +
  #scale_colour_manual(values = c("orange", "darkred", "darkgreen", "darkblue"  ))

LM_S <- filter(LM, EFmetric == "Stability") # subset for and plot EF: Stability

G3 <- ggplot(BEF_long, aes(x = DIVval, y = Stability, shape = Lake))+
    geom_point()+
  stat_smooth(method = "lm", se = FALSE, linetype = "solid", size = 0.5, colour = "black")+
  geom_rect( data = LM_S, aes(x=NULL,y=NULL, xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
             fill = "white", alpha = 0.8) +
  geom_text(data = LM_S, aes(x = x, y = y1, label = paste("r2.", signif(r.squared, 1), sep = " : " )), colour = "black", size = 2)+
  geom_text(data = LM_S, aes(x = x, y = y2, label = paste("p", signif(p.value, 1), sep = " : " )), colour = "black", size = 2)+
    facet_grid(Lake ~ DIVmetric, scales = "free")+
    theme_bw()+
    labs(title = "temporal stability of cell yield",
         y = "stability of cell yield",
         x = "")+
    theme(legend.position = "none",
          axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5),
          plot.margin=unit(c(1,1,0,0.5), "cm"))#+
    #scale_colour_manual(values = c("orange", "darkred", "darkgreen", "darkblue"  ))

LM_DIN <- filter(LM, EFmetric == "DIN") # subset for and plot EF: DIN
  
G4 <- ggplot(BEF_long, aes(x = DIVval, y = DIN, shape = Lake))+
    geom_point()+
  stat_smooth(method = "lm", se = FALSE, linetype = "solid", size = 0.5, colour = "black")+
   geom_rect( data = LM_DIN, aes(x=NULL,y=NULL, xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
             fill = "white", alpha = 0.8) +
  geom_text(data = LM_DIN, aes(x = x, y = y1, label = paste("r2.", signif(r.squared, 1), sep = " : " )), colour = "black", size = 2)+
  geom_text(data = LM_DIN, aes(x = x, y = y2, label = paste("p", signif(p.value, 1), sep = " : " )), colour = "black", size = 2)+
    facet_grid(Lake ~ DIVmetric, scales = "free")+
    theme_bw()+
    labs(title = "depletion of dissolved organic nitrogene",
         y = "-1 * standardized DIN concentration",
         x = "value of diveristy metric")+
    theme(legend.position = "none",
          axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5),
          plot.margin=unit(c(0,0.5,1,1), "cm")) #+
    #scale_colour_manual(values = c("orange", "darkred", "darkgreen", "darkblue"  ))


########## calculate Multifunctionality at 75% threshold ###############
  
MF_DF <- BEF_long %>% # calculate how many functions are over threshold for given diversity value
  group_by(DIVmetric, Lake) %>% 
  do( getFuncMaxed(., vars = c("DIN", "Stability", "maxCells"), thresh = 0.75, prepend = "DIVval", maxN = 2))

LM_MF <- MF_DF %>% #regress each div metric against the number of EF of threshold. extract p-value and r-square
  group_by(Lake, DIVmetric) %>% 
  do( glance( lm(scale(funcMaxed) ~ scale(DIVval), .))) %>% 
  select(Lake, DIVmetric, r.squared, p.value)

LM_MF <- LM %>% ungroup %>% select(Lake, DIVmetric, x, xmin, xmax) %>% distinct %>% left_join(LM_MF ,.) # add x-values for adding of regression stat

G5 <- ggplot(MF_DF, aes(x = DIVval, y = funcMaxed, shape = Lake))+ # plot diversity ~multifunctionality
    geom_point() +
  stat_smooth(method = "lm", se = FALSE, linetype = "solid", colour = "black", size = 0.5)+
   geom_rect( data = LM_MF, aes(x=NULL,y=NULL, xmin = xmin, xmax = xmax, ymin = 1.8, ymax = 2.7),
             fill = "white", alpha = 0.8) +
  geom_text(data = LM_MF, aes(x = x, y = 2.5, label = paste("r2.", signif(r.squared, 1), sep = " : " )), colour = "black", size = 2)+
  geom_text(data = LM_MF, aes(x = x, y = 2, label = paste("p", signif(p.value, 1), sep = " : " )), colour = "black", size = 2)+
    facet_grid(Lake ~ DIVmetric, scales="free_x")+
    theme_bw()+
    labs(title = "multifunctionality at 75 % threshold",
         y = "number of function > 75% of maximum",
         x = "value of diveristy metric")+
    theme(legend.position = "none",
          axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5),
          plot.margin=unit(c(0,1,1,0.5), "cm"))#+
    #scale_colour_manual(values = c("orange", "darkred", "darkgreen", "darkblue"  ))

G2
G3
G4
G5
  
pdf("figures/Figure_3.pdf", height = 10, width = 10)
grid.arrange(G2, G3, G4, G5, ncol = 2, top = "Pearson correlation")
dev.off()
#ggsave(plot = Div_G, "figures/Figure_3.pdf", height = 10, width = 10)

```

It is visible from the scatterplots that there doesn't seem to be a strong correlation between either of the diveristy metrics and either of the metrics of ecosystem functioning. 

the statics that we present in the plots are linear correlations of the form Ecosystem Function ~ Diversity Metric. As assumption of normality are however not necessarily met, we present aslo the rank correlations:


```{r, echo = FALSE}

LM_rank <- BEF_long2 %>%
  group_by(Lake, DIVmetric, EFmetric) %>% 
  do( glance( cor.test( .$EFval , .$DIVval, method = "spearman"))) %>% 
  select(Lake, DIVmetric, EFmetric, estimate, p.value)

LM_rank <- select(LM, -r.squared, -p.value) %>% 
  left_join(., LM_rank)


LMmC_R <- filter(LM_rank, EFmetric == "maxCells")
  
G6 <-  ggplot(BEF_long, aes(x = DIVval, y = maxCells, shape = Lake))+
  geom_point()+
  stat_smooth(method = "lm", se = FALSE, linetype = "solid", size = 0.5, colour = "black")+
  geom_rect( data = LMmC_R, aes(x=NULL,y=NULL, xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
             fill = "white", alpha = 0.8) +
  geom_text(data = LMmC_R, aes(x = x, y = y1, label = paste("r2.", signif(estimate, 1), sep = " : " )), colour = "black", size = 2)+
  geom_text(data = LMmC_R, aes(x = x, y = y2, label = paste("p", signif(p.value, 1), sep = " : " )), colour = "black", size = 2)+
  facet_grid(Lake ~ DIVmetric, scales = "free")+
  theme_bw()+
  labs(title = "maximum cell yield",
       y = "maximum Cell yield",
       x = "")+
  theme(legend.position = "none",
       axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5),
       plot.margin = unit(c(1,0.5,0,1), "cm"))+
  scale_y_continuous(label=scientific_10)#+
  #scale_colour_manual(values = c("orange", "darkred", "darkgreen", "darkblue"  ))

LM_S_R <- filter(LM_rank, EFmetric == "Stability")

G7 <- ggplot(BEF_long, aes(x = DIVval, y = Stability, shape = Lake))+
    geom_point()+
  stat_smooth(method = "lm", se = FALSE, linetype = "solid", size = 0.5, colour = "black")+
  geom_rect( data = LM_S_R, aes(x=NULL,y=NULL, xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
             fill = "white", alpha = 0.8) +
  geom_text(data = LM_S_R, aes(x = x, y = y1, label = paste("r2.", signif(estimate, 1), sep = " : " )), colour = "black", size = 2)+
  geom_text(data = LM_S_R, aes(x = x, y = y2, label = paste("p", signif(p.value, 1), sep = " : " )), colour = "black", size = 2)+
    facet_grid(Lake ~ DIVmetric, scales = "free")+
    theme_bw()+
    labs(title = "temporal stability of cell yield",
         y = "stability of cell yield",
         x = "")+
    theme(legend.position = "none",
          axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5),
          plot.margin=unit(c(1,1,0,0.5), "cm"))

LM_DIN_R <- filter(LM_rank, EFmetric == "DIN")
  
G8 <- ggplot(BEF_long, aes(x = DIVval, y = DIN, shape = Lake))+
    geom_point()+
  stat_smooth(method = "lm", se = FALSE, linetype = "solid", size = 0.5, colour = "black")+
   geom_rect( data = LM_DIN_R, aes(x=NULL,y=NULL, xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
             fill = "white", alpha = 0.8) +
  geom_text(data = LM_DIN_R, aes(x = x, y = y1, label = paste("r2.", signif(estimate, 1), sep = " : " )), colour = "black", size = 2)+
  geom_text(data = LM_DIN_R, aes(x = x, y = y2, label = paste("p", signif(p.value, 1), sep = " : " )), colour = "black", size = 2)+
    facet_grid(Lake ~ DIVmetric, scales = "free")+
    theme_bw()+
    labs(title = "depletion of dissolved organic nitrogene",
         y = "-1 * standardized DIN concentration",
         x = "value of diveristy metric")+
    theme(legend.position = "none",
          axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5),
          plot.margin=unit(c(0,0.5,1,1), "cm")) #+
    #scale_colour_manual(values = c("orange", "darkred", "darkgreen", "darkblue"  ))

LM_MF_rank <- MF_DF %>% #regress each div metric against the number of EF of threshold. extract p-value and r-square
  group_by(Lake, DIVmetric) %>% 
  do( glance( cor.test(.$funcMaxed , .$DIVval, method = "spearman"))) %>% 
  select(Lake, DIVmetric, estimate, p.value)

LM_MF_R <- LM_rank %>% ungroup %>% select(Lake, DIVmetric, x, xmin, xmax) %>% distinct %>% left_join(LM_MF_rank ,.) # add x-values for adding of



G9 <- ggplot(MF_DF, aes(x = DIVval, y = funcMaxed, shape = Lake))+ # plot diversity ~multifunctionality
    geom_point() +
  stat_smooth(method = "lm", se = FALSE, linetype = "solid", colour = "black", size = 0.5)+
   geom_rect( data = LM_MF_R, aes(x=NULL,y=NULL, xmin = xmin, xmax = xmax, ymin = 1.8, ymax = 2.7),
             fill = "white", alpha = 0.8) +
  geom_text(data = LM_MF_R, aes(x = x, y = 2.5, label = paste("r2.", signif(estimate, 1), sep = " : " )), colour = "black", size = 2)+
  geom_text(data = LM_MF_R, aes(x = x, y = 2, label = paste("p", signif(p.value, 1), sep = " : " )), colour = "black", size = 2)+
    facet_grid(Lake ~ DIVmetric, scales="free_x")+
    theme_bw()+
    labs(title = "multifunctionality at 75 % threshold",
         y = "number of function > 75% of maximum",
         x = "value of diveristy metric")+
    theme(legend.position = "none",
          axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5),
          plot.margin=unit(c(0,1,1,0.5), "cm"))


G6
G7
G8
G9
  
pdf("figures/Figure_S_2.pdf", height = 10, width = 10)
grid.arrange(G6, G7, G8, G9, ncol = 2, top = "Spearman rank correlation")
dev.off()


```

